The Multi-Stencil Language: orchestrating stencils with a mesh-agnostic DSL

### 2016-06-27 ARTICLES/OTHER/JLESC/slides.tex
JLESC presentation

\usepackage[absolute,overlay]{textpos}
\usepackage{tcolorbox}

@@

%-------------------------------------------------------------------
\title[From DSL to HPC Component-Based Runtime: A Multi-Stencil DSL Case Study]{From DSL to HPC Component-Based Runtime: A Multi-Stencil DSL Case Study}
\author[Julien Bigot (CEA), Hélène Coullon (INRIA), Christian Perez (INRIA)]{Julien Bigot, \underline{Hélène Coullon}, Christian Perez}
\institute[INRIA]{INRIA team Avalon\\Maison de la simulation (CEA)}
\date{WOLFHPC 2015 - $16^{th}$ November 2015}
%-------------------------------------------------------------------

@@

%---------------
%---------------
\begin{withoutheadline}

@@

% INTRODUCTION / MOTIVATION
%-------------------------------------------------------------
\begin{frame}
\frametitle{Motivation} % HPC DSL
\begin{block}{+ Domain Specific Languages}
\begin{itemize}
\item Separation of concerns (domain/implementation)
\item Easy language for the user
\item Implicit optimizations
\item Implicit parallelization
\end{itemize}
\end{block}
\begin{alertblock}{- Domain Specific Languages}
\begin{itemize}
\item Difficulties deported to the DSL designer
\begin{itemize}
\item Low level high performance programming
\item Maintainability and portability
\end{itemize}
\item As many DSLs as domains
\begin{itemize}
\item DSL composition ?
\end{itemize}
\end{itemize}
\end{alertblock}
\end{frame}
%-------------------------------------------------------------
\begin{frame}
\frametitle{Motivation} % components
\begin{block}{Component models}
\begin{itemize}
\item Divide an application into several independent black boxes
\item Each component defines its interactions with outer world
\item Application = Assembly of components
\end{itemize}
\end{block}
\begin{block}{+ Component models}
\begin{itemize}
\item Maintainability through separation of concerns
\item Code-reuse and productivity
\item Dynamic assembly of components
\end{itemize}
\end{block}
\end{frame}
%-------------------------------------------------------------
\begin{frame}
\frametitle{Motivation} % components
\begin{tcolorbox}[width=\textwidth,title={Motivation}]
What if a DSL produces a component-based runtime ?
\begin{itemize}
\item Is it feasible ?
\item Is it efficient ?
\item Does it improve issues of DSLs ? (maintainability etc.)
\end{itemize}
\end{tcolorbox}

\begin{block}{Contributions}
\begin{itemize}
\item The Multi-Stencil Language (MSL)
\item A first component-based back end
\item No overheads introduced
\end{itemize}
\end{block}

\end{frame}
%-------------------------------------------------------------

% presentation of the table of contents
%-------------------------------------------------------------
\begin{frame}
\frametitle{Table of contents}
    \tableofcontents[hideallsubsections]
\end{frame}
%-------------------------------------------------------------

\end{withoutheadline}
%---------------
%---------------

%-------------------------------------------------------------
\section{Multi-Stencil Applications}
% related work, multi-stencil problem, language
%-------------------------------------------------------------
%----------------
\subsection{Multi-Stencil Applications}
%----------------
\begin{frame}
\frametitle{Numerical simulation = Multi-Stencil application}
\begin{center}
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}
\def \n {5}
\def \radius {3cm}
\def \margin {8} % margin in angles, depends on the radius

  \node[purp2,align=center] (edp) at ({360/\n * (2 - 1)}:\radius) {Partial Differential Equations\\\scriptsize $\frac{\partial u(x,y,t)}{\partial t} = \frac{\partial^2 u(x,y,t)}{\partial x^2} + \frac{\partial^2 u(x,y,t)}{\partial y^2}$};
  \node[progm,align=center,right of=edp,node distance=6cm] (cond) {\scriptsize + specific behavior for\\\scriptsize boundary conditions};
  \path[line,dashed] (edp) -- (cond);
  \node[purp,align=center] (maillage) at ({360/\n * (1 - 1)}:\radius) {Mesh\\Time iterations};
  \path[->] (maillage) edge[<-,thin,shorten <=2pt,shorten >=2pt, bend right=40]  node [fill=white,align=center] {\scriptsize Time and space\\\scriptsize discretization} (edp.east);
  \path[] (edp.west) edge[thin,shorten <=2pt,shorten >=2pt, bend right=90]  node [draw,circle,fill=white] (plus) {\scriptsize +} (maillage.west);
  \node[purp,align=center, below of=plus, node distance=3.5cm] (sche) {Explicit\\numerical schemes\\\textcolor{red}{stencils}};
  \node[progm,align=center,right of=sche,node distance=5.5cm] (expl) {\scriptsize Step by step approximation of the phenomena};
  \path[line,dashed] (sche) -- (expl);
  \path[line] (plus) -- node [fill=white,align=center] {\scriptsize Numerical methods\\\scriptsize Finite difference/volume/elment} (sche);
\end{tikzpicture}
}
\end{center}
\end{frame}
%----------------

%----------------
\begin{frame}
\frametitle{Time and Mesh}
\begin{block}{Time}
At each time iteration of the simulation are applied the \emph{computation kernels} of the application.
\end{block}
\begin{block}{Mesh}
\begin{itemize}
\item A Mesh is a connected undirected graph $\mathcal{M}=(V,E)$ without bridges
\item Mesh entities are a subset of $V \cup E$
\end{itemize}
\end{block}
\begin{center}
\resizebox{6cm}{!}{\includegraphics{./images/mesh.pdf}}
\end{center}
\end{frame}
%----------------

%----------------
\begin{frame}
\frametitle{Data and Computation Kernels}
\begin{block}{Data}
Data is a set of numerical values, each one attached to a given mesh entity
\end{block}
\begin{block}{Computation kernel}
\begin{itemize}
\item Set of data read for the computation
\begin{itemize}
\item Each one associated to a stencil shape
\end{itemize}
\item Data written by the computation
\item A numerical expression
\item A computation domain
\begin{itemize}
\item Subset of mesh entities
\end{itemize}
\end{itemize}
\end{block}
\end{frame}
%----------------

%----------------
% \begin{frame}
% \frametitle{Computation kernel: Examples}
% \begin{center}
% \resizebox{6cm}{!}{\includegraphics{./images/mesh.pdf}}
% \hspace{10pt}
% \resizebox{5cm}{!}{\includegraphics{./images/stencil1.pdf}}
% \end{center}
% Compute $A$ using $\{(B,n1)\}$ on the computation domain called $dc1$ with the numerical expression
% \tiny
% \begin{equation*}
% A(x,y)=B(x+1,y)+B(x-1,y)+B(x,y+1)+B(x,y-1).
% \end{equation*}
% \end{frame}
%----------------

%----------------
% \begin{frame}
% \frametitle{Computation kernel: Example}
% \begin{center}
% \resizebox{6cm}{!}{\includegraphics{./images/mesh.pdf}}
% \hspace{10pt}
% \resizebox{5cm}{!}{\includegraphics{./images/stencil2.pdf}}
% \end{center}
% \end{frame}
%----------------

%----------------
\begin{frame}
\frametitle{Multi-Stencil application}
\begin{center}
$\mathcal{MSP}(T,\mathcal{M},\mathcal{E},\mathcal{D},\Delta,\Gamma)$
\end{center}
\begin{itemize}
\item $T$ the set of time iterations to tun the simulation
\item $\mathcal{M}$ the mesh of the simulation
\item $\mathcal{E}$ the set of mesh entities
\item $\mathcal{D}$ the set of computation domains
\item $\Delta$ the set of data
\item $\Gamma$ the set of computations
\end{itemize}
\center \textit{= the six sections of a \emph{Multi-Stencil Language} program !}
\end{frame}
%----------------

%-------------------------------------------------------------
\section{MSL Overview}
% take the same notations inside components + user
%-------------------------------------------------------------
\subsection{MSL Overview}

%----------------
\begin{frame}[fragile]
\frametitle{Example}
\begin{columns}
\begin{column}{.52\textwidth}
\resizebox{\textwidth}{!}{\includegraphics{./images/mesh.pdf}}
\hspace{10pt}
\resizebox{\textwidth}{!}{\includegraphics{./images/stencil2.pdf}}
\end{column}
\begin{column}{.48\textwidth}
$\mathcal{MSP}(T,\mathcal{M},\mathcal{E},\mathcal{D},\Delta,\Gamma)$\\

\medskip
\textit{\textcolor{blue}{blue} = MSL and black = user identifiers}

\medskip
\begin{lstlisting}[style=base]
@mesh:@ cart
@mesh entities:@ cell@,@edgex
@computation domains:@
  allcell @in@ cell
  alledgex @in@ edgex
@data:@
  A@,@cell
  C@,@edgex
@time:@500
@computations:@
  A@[@allcell@]@=&comp&@(@C@[@n1@]@@)@
\end{lstlisting}
\end{column}
\end{columns}
\end{frame}

%----------------
\begin{frame}[fragile]
\frametitle{Where is the code ?}
Keywords + identifiers ? That's it ? Where is the code ???
\end{frame}

%----------------
\begin{frame}[fragile]
\frametitle{Where is the code ?}
\begin{block}{Data parallelism}
Whatever the mesh is and the exact computations are:
\begin{itemize}
\item mesh and data are splitted among resources
\item synchronizations detected from \textbf{computation dependencies}
\end{itemize}
\end{block}

\begin{block}{Mid-grain task parallelism}
Task = complete computation\\
Whatever the mesh is and the exact computations are:
\begin{itemize}
\item task dependencies detected from \textbf{computation dependencies}
\end{itemize}
\end{block}

\begin{alertblock}{Fine grain optimizations and parallelizations}
Stencil compilers already exist !
\end{alertblock}

\end{frame}

%----------------
\begin{frame}[fragile]
\frametitle{Where is the code ?}
\begin{block}{Data parallelism}
Whatever the mesh is and the exact computations are:
\begin{itemize}
\item mesh and data are splitted among resources
\item synchronizations detected from computation dependencies
\end{itemize}
\end{block}

\begin{block}{Mid-grain task parallelism}
Task = complete computation\\
Whatever the mesh is and the exact computations are:
\begin{itemize}
\item task dependencies detected from computation dependencies
\end{itemize}
\end{block}

\begin{alertblock}{Fine grain optimizations and parallelizations}
Stencil compilers already exist !
\end{alertblock}

\begin{textblock*}{0.8\textwidth}(20mm,32mm)
\begin{tcolorbox}[width=\textwidth,title={MSL = computation dependencies}]
  \begin{itemize}  
  \item produces a component based parallel scheduling of the simulation
  \item it is agnostic from the mesh
  \item it is agnostic from the actual numerical code
  \end{itemize}
\end{tcolorbox} 
\end{textblock*}
% based parallel scheduling of the simulation\\ It is agnostic from the mesh\\it is agnostic from the actual numerical code

\end{frame}

% %----------------
% \begin{frame}[fragile]
% \frametitle{MSL Example}
% \begin{columns}
% \begin{column}{.52\textwidth}
% \begin{lstlisting}[style=base]
% @mesh:@ cart
% @mesh entities:@ cell,edgex,edgey
% @computation domains:@
%   allcell in cell
%   alledgex in edgex
%   alledgey in edgey
%   part1edgex in edgex
%   part2edgex in edgex
% @data:@
%   a,cell
%   b,cell
%   c,edgex
%   d,edgex
%   e,edgey
% \end{lstlisting}
% \end{column}
% \begin{column}{.48\textwidth}
% \begin{lstlisting}[style=base]
%   f,cell
%   g,edgey
%   h,edgex
%   i,cell
%   j,edgex
% @time:@500
% @computations:@
%   b[allcell]=&c0&(a)
%   c[alledgex]=&c1&(b[n1])
%   d[alledgex]=&c2&(c)
%   e[alledgey]=&c3&(c)
%   f[allcell]=&c4&(d[n1])
%   g[alledgey]=&c5&(e)
%   h[alledgex]=&c6&(f)
%   i[allcell]=&c7&(g,h)
%   j[partedgex]=&c8&(i[n1])
% \end{lstlisting}
% \end{column}
% \end{columns}
% \end{frame}
% %----------------

% %----------------
% \begin{frame}
% \frametitle{Multi-Stencil Language}
% \begin{alertblock}{MSL is not}
% \begin{itemize}
% \item a new stencil optimizer/compiler
% \item a new distributed data structure
% \end{itemize}
% \end{alertblock}
% \begin{block}{MSL is}
% \begin{itemize}
% \item a high-level language for multi-stencil simulations
% \item agnostic from the type of mesh used (data structure)
% \end{itemize}
% \end{block}
% \begin{center}
% \textit{MSL produces an empty\\component-based parallel scheduling of the simulation}
% \end{center}
% \end{frame}
% %----------------

%----------------
\begin{frame}
\frametitle{Related Work}
\begin{block}{Complementary work}
\begin{itemize}
\item Distributed data structures: SkelGIS, Global Arrays
\item Stencil DSLs (on grids): Pochoir, PATUS
\item Stencil DSLs (on unstructured meshes): OP2, Liszt
\end{itemize}
\end{block}
\begin{alertblock}{Similar work}
\begin{itemize}
\item Pipeline of stencil computations for image processing: Halide
\begin{itemize}
\item On grids (image), different abstraction level
\end{itemize}
\item DSL to component-based runtime ?
\end{itemize}
\end{alertblock}
\end{frame}
%----------------
%-------------------------------------------------------------

% %----------------
% \begin{frame}
% \frametitle{MSL to Component-based runtime}
% \begin{block}{Ready-to-fill parallel scheduling: mid-grain parallelism}
% \begin{itemize}
% \item Data parallelism
% \begin{itemize}
% \item External distributed data structure
% \item Automatic detection of synchronizations
% \end{itemize}
% \item Task parallelism
% \begin{itemize}
% \item Compile a static scheduling of computation kernels
% \end{itemize}
% \end{itemize}
% \end{block}
% The fine grain parallelism is left to other languages: 
% \begin{itemize}
% \item OpenMP in the kernels
% \item Kernels generated by stencil compilers for CPU or GPU (Pochoir, Liszt etc.)
% \end{itemize}
% \end{frame}
% %----------------

%----------------
% \begin{frame}
% \frametitle{MSL to Component-based runtime}
% \begin{block}{Empty component-based parallel scheduling of the simulation}
% \begin{itemize}
% \item Data parallelism
% \begin{itemize}
% \item Use of an external distributed data structure
% \item Automatic detection of synchronizations
% \end{itemize}
% \item Mid-grain task parallelism
% \begin{itemize}
% \item A task is a complete computation (space loop)
% \end{itemize}
% \end{itemize}
% \end{block}
% The fine grain optimizations/parallelizations are left to other languages: 
% \begin{itemize}
% \item OpenMP in the kernels
% \item Kernels generated by stencil compilers (Pochoir, Liszt etc.)
% \end{itemize}
% \end{frame}
%----------------

%----------------
\begin{frame}
\frametitle{MSL to Component-based runtime}
\begin{center}
$\mathcal{MSP}(T,\mathcal{M},\mathcal{E},\mathcal{D},\Delta,\Gamma)$\\

\bigskip
\begin{tikzpicture}[remember picture]
  \node[rectangle,rounded corners=3pt,thick,inner sep=5pt,draw=black,dotted] (proc) {
    \begin{tikzpicture}[shorten >=1pt, node distance=2cm, on grid, auto]
     \node[component,solid,draw=red] (D) at (0,0) {$Driver$};
     \node[provide,scale=0.5,solid] (Dp) at (-1,0) {};
     \node[solid] (Ds) at (-1.5,0) {start};
     \node[use,scale=0.5,solid,right=1.5cm of D] (Du1) {};
     \node[use,scale=0.5,solid,below=1.75cm of D] (Du2) {};
     \node[use,scale=0.5,solid,below=0.8cm of Du1] (Du3) {$m$};

     \node[provide,scale=0.5,solid,below=0.15 of Du2] (Tp) {};
     \node[component,solid,below=1.6cm of Tp] (T) {$Time$};
     \node[use,scale=0.5,solid,right=1cm of T] (Tu) {};
     \node[solid,left=0.8cm of T] (tt) {$T$};

     \node[provide,scale=0.5,solid,right=0.15 of Tu] (Cp) {};
     \node[component,dashed,right=2cm of Cp] (C) {$Computations$};
     \node[use,scale=0.5,solid,above=0.8cm of C] (Cu) {};
     \node[solid,right=0.2cm of Cu] (star) {$*$};
     \node[solid,right=1.5cm of C] (gamma) {$\Gamma$};

     \node[provide,scale=0.5,solid,below right=0.2 of Du3] (Datap1) {};
     \node[provide,scale=0.5,solid,above=0.15 of Cu] (Datap2) {};
     \node[component,solid,double,above=0.8cm of Datap2] (Data) {$Data$};
     \node[use,scale=0.5,solid,above=0.8cm of Data] (Datau) {};
     \node[solid,left=1cm of Data] (delta) {$\Delta,\mathcal{D}$};
     %\node[mpi,scale=0.8,rounded corners=0pt,solid,right=1cm of Data] (mpidata) {};
     %\node[solid,above=0.2cm of mpidata] (star2) {$*$};

     \node[provide,scale=0.5,solid,right=0.15 of Du1] (DDSp1) {};
     \node[provide,scale=0.5,solid,above=0.15 of Datau] (DDSp2) {};
     \node[component,solid,above=0.8cm of DDSp2] (DDS) {$DDS$};
     \node[solid,right=1.2cm of DDS] (m) {$\mathcal{M},\mathcal{E}$};

    \path[-]
      (Dp) edge [solid] node {} (D)
      (D) edge [solid] node {} (Du1)
          edge [solid] node {} (Du2)
          edge [solid] node {} (Du3)
      (DDSp1) edge [solid] node {} (DDS)
      (Tp) edge [solid] node {} (T)
      (T)  edge [solid] node {} (Tu)
      (Cp) edge [solid] node {} (C)
      (C) edge [solid] node {} (Cu)
      (Datap2) edge [solid] node {} (Data)
      (Datap1) edge [solid] node {} (Data)
      (Data) edge [solid] node {} (Datau)
      (DDSp2) edge [solid] node {} (DDS);
    \end{tikzpicture}
    };
    \node[mpi,scale=0.8,rounded corners=0pt,solid,right=0.5cm of proc] (mpidata) {};
    \node[solid,above=0.1cm of mpidata] (star2) {$*$};
    \node[solid,below=2.5cm of proc.west,anchor=west] (core) {\textit{\tiny{Duplicated on each processor/core}}};
    \path[-]
      (Data.east) edge node {} (mpidata);
  \end{tikzpicture}
\end{center}
\end{frame}
%-------------------------------------------------------------

%----------------
\begin{frame}
\frametitle{MSL to Component-based runtime}
\begin{center}
$\mathcal{MSP}(T,\mathcal{M},\mathcal{E},\mathcal{D},\Delta,\Gamma)$\\

\bigskip
\begin{tikzpicture}[remember picture]
  \node[rectangle,rounded corners=3pt,thick,inner sep=5pt,draw=red,dotted] (proc) {
    \begin{tikzpicture}[shorten >=1pt, node distance=2cm, on grid, auto]
     \node[component,solid] (D) at (0,0) {$Driver$};
     \node[provide,scale=0.5,solid] (Dp) at (-1,0) {};
     \node[solid] (Ds) at (-1.5,0) {start};
     \node[use,scale=0.5,solid,right=1.5cm of D] (Du1) {};
     \node[use,scale=0.5,solid,below=1.75cm of D] (Du2) {};
     \node[use,scale=0.5,solid,below=0.8cm of Du1] (Du3) {$m$};

     \node[provide,scale=0.5,solid,below=0.15 of Du2] (Tp) {};
     \node[component,solid,below=1.6cm of Tp] (T) {$Time$};
     \node[use,scale=0.5,solid,right=1cm of T] (Tu) {};
     \node[solid,left=0.8cm of T] (tt) {$T$};

     \node[provide,scale=0.5,solid,right=0.15 of Tu] (Cp) {};
     \node[component,dashed,right=2cm of Cp] (C) {$Computations$};
     \node[use,scale=0.5,solid,above=0.8cm of C] (Cu) {};
     \node[solid,right=0.2cm of Cu] (star) {$*$};
     \node[solid,right=1.5cm of C] (gamma) {$\Gamma$};

     \node[provide,scale=0.5,solid,below right=0.2 of Du3] (Datap1) {};
     \node[provide,scale=0.5,solid,above=0.15 of Cu] (Datap2) {};
     \node[component,solid,draw=red,double,above=0.8cm of Datap2] (Data) {$Data$};
     \node[use,scale=0.5,solid,above=0.8cm of Data] (Datau) {};
     \node[solid,left=1cm of Data] (delta) {$\Delta,\mathcal{D}$};
     %\node[mpi,scale=0.8,rounded corners=0pt,solid,right=1cm of Data] (mpidata) {};
     %\node[solid,above=0.2cm of mpidata] (star2) {$*$};

     \node[provide,scale=0.5,solid,right=0.15 of Du1] (DDSp1) {};
     \node[provide,scale=0.5,solid,above=0.15 of Datau] (DDSp2) {};
     \node[component,draw=red,solid,above=0.8cm of DDSp2] (DDS) {$DDS$};
     \node[solid,right=1.2cm of DDS] (m) {$\mathcal{M},\mathcal{E}$};

    \path[-]
      (Dp) edge [solid] node {} (D)
      (D) edge [solid] node {} (Du1)
          edge [solid] node {} (Du2)
          edge [solid] node {} (Du3)
      (DDSp1) edge [solid] node {} (DDS)
      (Tp) edge [solid] node {} (T)
      (T)  edge [solid] node {} (Tu)
      (Cp) edge [solid] node {} (C)
      (C) edge [solid] node {} (Cu)
      (Datap2) edge [solid] node {} (Data)
      (Datap1) edge [solid] node {} (Data)
      (Data) edge [solid] node {} (Datau)
      (DDSp2) edge [solid] node {} (DDS);
    \end{tikzpicture}
    };
    \node[mpi,scale=0.8,rounded corners=0pt,solid,right=0.5cm of proc] (mpidata) {};
    \node[solid,above=0.1cm of mpidata] (star2) {$*$};
    \node[solid,below=2.5cm of proc.west,anchor=west] (core) {\textit{\tiny{Duplicated on each processor/core}}};
    \path[-]
      (Data.east) edge node {} (mpidata);
  \end{tikzpicture}
\end{center}
\end{frame}
%-------------------------------------------------------------

%----------------
\begin{frame}
\frametitle{MSL to Component-based runtime}
\begin{center}
$\mathcal{MSP}(T,\mathcal{M},\mathcal{E},\mathcal{D},\Delta,\Gamma)$\\

\bigskip
\begin{tikzpicture}[remember picture]
  \node[rectangle,rounded corners=3pt,thick,inner sep=5pt,draw=black,dotted] (proc) {
    \begin{tikzpicture}[shorten >=1pt, node distance=2cm, on grid, auto]
     \node[component,solid] (D) at (0,0) {$Driver$};
     \node[provide,scale=0.5,solid] (Dp) at (-1,0) {};
     \node[solid] (Ds) at (-1.5,0) {start};
     \node[use,scale=0.5,solid,right=1.5cm of D] (Du1) {};
     \node[use,scale=0.5,solid,below=1.75cm of D] (Du2) {};
     \node[use,scale=0.5,solid,below=0.8cm of Du1] (Du3) {$m$};

     \node[provide,scale=0.5,solid,below=0.15 of Du2] (Tp) {};
     \node[component,solid,draw=red,below=1.6cm of Tp] (T) {$Time$};
     \node[use,scale=0.5,solid,right=1cm of T] (Tu) {};
     \node[solid,left=0.8cm of T] (tt) {$T$};

     \node[provide,scale=0.5,solid,right=0.15 of Tu] (Cp) {};
     \node[component,dashed,draw=red,right=2cm of Cp] (C) {$Computations$};
     \node[use,scale=0.5,solid,above=0.8cm of C] (Cu) {};
     \node[solid,right=0.2cm of Cu] (star) {$*$};
     \node[solid,right=1.5cm of C] (gamma) {$\Gamma$};

     \node[provide,scale=0.5,solid,below right=0.2 of Du3] (Datap1) {};
     \node[provide,scale=0.5,solid,above=0.15 of Cu] (Datap2) {};
     \node[component,solid,double,above=0.8cm of Datap2] (Data) {$Data$};
     \node[use,scale=0.5,solid,above=0.8cm of Data] (Datau) {};
     \node[solid,left=1cm of Data] (delta) {$\Delta,\mathcal{D}$};
     %\node[mpi,scale=0.8,rounded corners=0pt,solid,right=1cm of Data] (mpidata) {};
     %\node[solid,above=0.2cm of mpidata] (star2) {$*$};

     \node[provide,scale=0.5,solid,right=0.15 of Du1] (DDSp1) {};
     \node[provide,scale=0.5,solid,above=0.15 of Datau] (DDSp2) {};
     \node[component,solid,above=0.8cm of DDSp2] (DDS) {$DDS$};
     \node[solid,right=1.2cm of DDS] (m) {$\mathcal{M},\mathcal{E}$};

>>>>

@@

%-------------------------------------------------------------------
\title[The Multi-Stencil Language]{The Multi-Stencil Language: orchestrating stencils with a mesh-agnostic DSL}
\author[Hélène Coullon (INRIA), Julien Bigot (CEA), Christian Perez (INRIA)]{\underline{Hélène Coullon}, Julien Bigot, Christian Perez}
\institute[INRIA]{INRIA team Avalon\\Maison de la simulation (CEA)}
\date{$5{th}$ JLESC Workshop - 28th June 2016, Lyon}
%-------------------------------------------------------------------

@@

%---------------

@@

%-------------------------------------------------------------
% INTRODUCTION / MOTIVATION
%-------------------------------------------------------------
\section{Introduction}
%-------------------------------------------------------------
\begin{frame}
\frametitle{Domain Specific Languages} % HPC DSL
\begin{block}{+ Domain Specific Languages}
Separation of concerns between the domain and its efficient implementation
\begin{itemize}
\item Easy language for the end-user
\item Implicit optimizations
\item Implicit parallelization
\end{itemize}
\end{block}
But none or only a few DSLs used in production !
\end{frame}
%-------------------------------------------------------------
\begin{frame}
\frametitle{Domain Specific Languages} % HPC DSL
\begin{alertblock}{- Domain Specific Languages}
\begin{itemize}
\item Choose a domain and a language more or less specific
\item Difficult to combine DSLs (interoperability)
\begin{itemize}
\item semantic / compilation / back-end codes
\end{itemize}
\item Difficult to maintain
\item Exascale applications = many specific domains and interactions
\end{itemize}
\end{alertblock}
\end{frame}
%-------------------------------------------------------------
\begin{frame}
\frametitle{How to improve it ?} % HPC DSL
\begin{itemize}
\item Choose the good abstraction level to
\begin{itemize}
\item stay efficient
\item be used in many domains (find common meta-domains)
\end{itemize}
\item Choose good programming models for
\begin{itemize}
\item the compiler programming
\item the back-end
\end{itemize}
\item Try to reuse existing DSLs (compilation process or back-end codes)
\end{itemize}
\textit{Delite (Scala)\\
Composition and Reuse with Compiled Domain-Specific Languages\\
ECOOP'13}
\end{frame}
%-------------------------------------------------------------

%-------------------------------------------------------------
\section{Multi-Stencil Language : Yet another DSL for stencils !}
%-------------------------------------------------------------
\begin{frame}
\frametitle{Multi-Stencil application}
\begin{center}
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}
\def \n {5}
\def \radius {3cm}
\def \margin {8} % margin in angles, depends on the radius

  \node[purp2,align=center] (edp) at ({360/\n * (2 - 1)}:\radius) {Partial Differential Equations\\\scriptsize $\frac{\partial u(x,y,t)}{\partial t} = \frac{\partial^2 u(x,y,t)}{\partial x^2} + \frac{\partial^2 u(x,y,t)}{\partial y^2}$};
  \node[progm,align=center,right of=edp,node distance=6cm] (cond) {\scriptsize + specific behavior for\\\scriptsize boundary conditions};
  \path[line,dashed] (edp) -- (cond);
  \node[purp,align=center] (maillage) at ({360/\n * (1 - 1)}:\radius) {Mesh\\Time iterations};
  \path[->] (maillage) edge[<-,thin,shorten <=2pt,shorten >=2pt, bend right=40]  node [fill=white,align=center] {\scriptsize Time and space\\\scriptsize discretization} (edp.east);
  \path[] (edp.west) edge[thin,shorten <=2pt,shorten >=2pt, bend right=90]  node [draw,circle,fill=white] (plus) {\scriptsize +} (maillage.west);
  \node[purp,align=center, below of=plus, node distance=3.5cm] (sche) {Explicit\\numerical schemes\\\textcolor{red}{stencils}};
  \node[progm, text width=11em,align=center,right of=sche,node distance=5.5cm] (expl) {\scriptsize Iterative neighboorhood approximation of the phenomena};
  \path[line,dashed] (sche) -- (expl);
  \path[line] (plus) -- node [fill=white,align=center] {\scriptsize Numerical methods\\\scriptsize Finite difference/volume/elment} (sche);
\end{tikzpicture}
}
\end{center}
\end{frame}
%----------------
\begin{frame}
\frametitle{Multi-Stencil Language : Yet another DSL for stencils !}
\begin{block}{Inputs}
follow a descriptive grammar to simply declare
\begin{itemize}
\item the sequential order of computations
\item what is read and written by each computation
\item whith or without a neighborhood access (stencil)
\end{itemize}
\end{block}
\begin{block}{Outputs}
\begin{itemize}
\item parallel pattern of the multi-stencil application
\item hybrid parallelism included: data and task parallelism
\item choice of parallel languages (MPI, OpenMP, StarPU, SkelGIS, Global Arrays etc.)
\item end-users have to fill computation functions into this pattern
\end{itemize}
\end{block}
\end{frame}
%----------------
\begin{frame}
\frametitle{Multi-Stencil Language : Yet another DSL for stencils !}
\begin{itemize}
\item Choosen abstraction level
\begin{itemize}
\item mesh-agnostic description language
\begin{itemize}
\item should be usable for unstructured, Cartesian, curvilinear meshes
\end{itemize}
\item description concerns splited from implementation concerns
\begin{itemize}
\item usable with many different parallel libraries or languages back-ends
\end{itemize}
\end{itemize}
\item Choosen programming models for back-end
\begin{itemize}
\item Component programming model (maintainability, portability and composability)
\item Task scheduling model (portability and efficiency)
\end{itemize}
\item Reuse of existing languages and DSLs
\begin{itemize}
\item Use of a Distributed Data Structure DSL (MPI) : SkelGIS
\item MPI and OpenMP (4.x)
\end{itemize}
\end{itemize}
\end{frame}
%----------------
\begin{frame}
\frametitle{MSL to Component-based runtime}
\begin{center}
\begin{tikzpicture}[remember picture]
  \node[rectangle,rounded corners=3pt,thick,inner sep=5pt,draw=black,dotted] (proc) {
    \begin{tikzpicture}[shorten >=1pt, node distance=2cm, on grid, auto]
     \node[component,solid] (D) at (0,0) {$Driver$};
     \node[provide,scale=0.5,solid] (Dp) at (-1,0) {};
     \node[solid] (Ds) at (-1.5,0) {start};
     \node[use,scale=0.5,solid,right=1.5cm of D] (Du1) {};
     \node[use,scale=0.5,solid,below=1.75cm of D] (Du2) {};
     \node[use,scale=0.5,solid,below=0.8cm of Du1] (Du3) {$m$};

     \node[provide,scale=0.5,solid,below=0.15 of Du2] (Tp) {};
     \node[component,solid,below=1.6cm of Tp] (T) {$Time$};
     \node[use,scale=0.5,solid,right=1cm of T] (Tu) {};

     \node[provide,scale=0.5,solid,right=0.15 of Tu] (Sp) {};
     \node[component,solid,right=1.5cm of Sp] (S) {$Scheduling$};
     \node[use,scale=0.5,solid,right=1.5cm of S] (Su) {};

     \node[provide,scale=0.5,solid,right=0.15 of Su] (Cp) {};
     \node[component,solid,double,right=2cm of Cp] (C) {$Kernel$};
     \node[use,scale=0.5,solid,above=0.8cm of C] (Cu) {};

     \node[provide,scale=0.5,solid,below right=0.2 of Du3] (Datap1) {};
     \node[provide,scale=0.5,solid,above=0.15 of Cu] (Datap2) {};
     \node[component,solid,double,above=0.8cm of Datap2] (Data) {$Data$};
     \node[use,scale=0.5,solid,above=0.8cm of Data] (Datau) {};

     \node[provide,scale=0.5,solid,right=0.15 of Du1] (DDSp1) {};
     \node[provide,scale=0.5,solid,above=0.15 of Datau] (DDSp2) {};
     \node[component,solid,above=0.8cm of DDSp2] (DDS) {$DDS$};

    \path[-]
      (Dp) edge [solid] node {} (D)
      (D) edge [solid] node {} (Du1)
          edge [solid] node {} (Du2)
          edge [solid] node {} (Du3)
      (DDSp1) edge [solid] node {} (DDS)
      (Tp) edge [solid] node {} (T)
      (T)  edge [solid] node {} (Tu)
      (Sp) edge [solid] node {} (S)
      (S) edge [solid] node {} (Su)
      (Datap2) edge [solid] node {} (Data)
      (Datap1) edge [solid] node {} (Data)
      (Data) edge [solid] node {} (Datau)
      (C) edge [solid] node {} (Cu)
      	  edge [solid] node {} (Cp)
      (DDSp2) edge [solid] node {} (DDS);
    \end{tikzpicture}
    };
    \node[mpi,scale=0.8,rounded corners=0pt,solid,right=0.5cm of proc] (mpidata) {};
    \node[solid,above=0.1cm of mpidata] (star2) {$*$};
    \node[solid,below=2.5cm of proc.west,anchor=west] (core) {\textit{\tiny{Duplicated on each processor/core}}};
    \path[-]
      (Data.east) edge node {} (mpidata);
  \end{tikzpicture}
\end{center}
\end{frame}
%-------------------------------------------------------------
\section{Few results}
%----------------
\begin{frame}
\frametitle{Shallow-water equations}
\begin{center}
Cartesian mesh / finite volume method\\Strong scaling 10k x 10k mesh with 1k iterations\\TGCC Curie, Thin nodes
\resizebox{7cm}{!}{\includegraphics{./images/median_strong.pdf}}
\end{center}
\end{frame}
%----------------
\begin{frame}
\frametitle{Shallow-water equations}
\begin{center}
Cartesian mesh / finite volume method\\Weak scaling 400 x 400 per process\\TGCC Curie, Thin nodes
\resizebox{7cm}{!}{\includegraphics{./images/median_weak.pdf}}
\end{center}
\end{frame}
%----------------
% \begin{frame}
% \frametitle{Multi-Stencil Language : Yet another DSL for stencils !}
% \textbf{Improvment with data+task scheduling}
% \begin{center}
% \resizebox{8cm}{!}{\includegraphics{./images/base_close_median.pdf}}
% \end{center}
% \end{frame}

%-------------------------------------------------------------
\begin{frame}
\frametitle{Work in progress and collaborations}
\begin{block}{Work in progress}
\begin{itemize}
\item different types of scheduling and back-end in MSL
\item combine MSL implementation choices to an energy-aware framework
\item new programming model back-end for MSL: combining component models and task models (Ph.D. Jérôme Richard)
\end{itemize}
\end{block}
\begin{block}{Collaborations}
\begin{itemize}
\item More application cases
\item Back-end programming model: component models and task scheduling (Components in OmpSs ?)
\item DSLs interoperability
\end{itemize}
\end{block}
\end{frame}
%-------------------------------------------------------------

% \begin{withoutheadline}
% \begin{frame}{}
% \begin{center}
% \huge Thank You !
% \end{center}
% \end{frame}
% \end{withoutheadline}
